Reformulating phylogenetic mixed models to improve flexibility and speed

### 2019-09-10 phyloglmm_ms.tex

We have presented the range of classical phylogenetic comparative methods (i.e. phylogenetic least squares, linear and mixed models) to fit models that incorporates different levels of phylogenetic variation.
Even within this scope, there is additional room for exploration we neglected, such as phylogenetic multivariate response models; non-Brownian evolutionary processes such as the Ornstein-Uhlenbeck (OU) model which accounts for both selection and drift processes \citep{butler2004phylogenetic}); Bayesian approaches \citep{hadfield2010general}; and variable-rate model, where evolutionary parameters vary across the phylogeny.
However, the simple approach we developed here offers an efficient way to handle phylogenetic comparative analysis for a wide range of univariate, Brownian-motion evolutionary models. 
This approach can in principle be combined flexibly with any platforms that supports independent latent variables such as Stan.
More importantly, this implementation in \pkg{lme4} and \pkg{glmmTMB} allows users to fit phylogenetic mixed models to the fullest (large data, unbalanced species observations, complex random effects) and explore new ideas.

@@

ML and BMB conceived the ideas and designed methodology; ML and BMB implimented the code in \pkg{lme4} and \pkg{glmmTMB}; ML ran all simulations; ML and BMB analysed the results; ML wrote the first draft of the manuscript. All authors contributed critically to the drafts and gave final approval for publication.

>>>>

We have presented a range of classical phylogenetic comparative methods (i.e. phylogenetic least squares, linear and mixed models) to fit models that incorporate different levels of phylogenetic variation.
Even within this scope, there is additional room for exploration, such as phylogenetic multivariate response models; non-Brownian evolutionary processes such as the Ornstein-Uhlenbeck (OU) model which accounts for both selection and drift processes \citep{butler2004phylogenetic}); Bayesian approaches \citep{hadfield2010general}; and variable-rate model, where evolutionary parameters vary across the phylogeny.
However, the simple approach we developed here offers an efficient way to handle phylogenetic comparative analysis for a wide range of univariate, Brownian-motion evolutionary models. 
This approach can in principle be combined flexibly with any platforms that supports independent latent variables such as Stan.
More importantly, this implementation in \pkg{lme4} and \pkg{glmmTMB} allows users to fit phylogenetic mixed models to the fullest (large data, unbalanced species observations, complex random effects) and explore new ideas.

@@

ML and BMB conceived the ideas and designed methodology; ML and BMB implemented the code in \pkg{lme4} and \pkg{glmmTMB}; ML ran all simulations; ML and BMB analysed the results; ML wrote the first draft of the manuscript. All authors contributed critically to the drafts and gave final approval for publication.
